---
title: "CSML1000-003-O-F19 - Group 2 - Assignment 2"
author: "Rajiv Kaushik, Yuming Cui, Madana Bolla, Pratik Chandwani, Konstantin Krassavine"
date: November 3, 2019
output: html_document
---

```{r warning = FALSE, echo=FALSE}
# Install packages
# install.packages('corrplot', dependencies = TRUE)
# install.packages('cluster', dependencies = TRUE)
# install.packages('factoextra', dependencies = TRUE)
# install.packages('reshape2', dependencies = TRUE)
# install.packages('tidyverse', dependencies = TRUE)
# install.packages('RColorBrewer', dependencies = TRUE)
# install.packages('scales', dependencies = TRUE)
# install.packages('fpc', dependencies = TRUE)
# install.packages('EnvStats', dependencies = TRUE)
# install.packages('NbClust', dependencies = TRUE)
# install.packages('dendextend', dependencies = TRUE)
# install.packages('ClusterR', dependencies = TRUE)
```

# Import libraries
```{r warning = FALSE, message=FALSE, warning=FALSE}
library(corrplot)
library(cluster)
library(factoextra)
library(reshape2)
library(tidyverse)
library(RColorBrewer)
library(scales)
library(fpc)
library(EnvStats)
library(NbClust)
library(dendextend)
library(gtools)
library(ClusterR)
library(dendextend)
```

## Introduction
The dataset is related to variety of wine variants. The goal is to model wine quality based on physicochemical tests.

## Objective
These datasets can be viewed as classification or regression tasks. The classes are ordered and not balanced (e.g. there are many more normal wines than excellent or poor ones). Outlier detection algorithms could be used to detect the few excellent or poor wines. Also, we are not sure if all input variables are relevant. So it could be interesting to test feature selection methods.


# Data Analysis
The data set contains 18 variables that involve wine realted attributes.

## Data Dictionary

Column Name             | Description 
------------------------|------------------------------------------------|
1 - fixed acidity       | fixed acidity of the wine                      |
2 - volatile acidity    | volatile acidity of the wine                   |
3 - citric acid         | citric acid in wine observation                |
4 - residual sugar      | residual sugar in wine                         |
5 - chlorides           | cholorides in wine                             |
6 - free sulfur dioxide | free sulfur dioxide                            |
7 - total sulfur dioxide| total sulfur dioxide                           |
8 - density             | Wine density                                   |
9 - pH                  | Wine pH value                                  |
10 - sulphates          | sulphates in wine                              |
11 - alcohol            | Wine alchocoal level                           |
12 - #add rowid (0 to 10)  | wien quality based on based on sensory data)   |
13 - country            | country origin of wine                         |
14 - province           | province origin of wine                        |
15 - region_1           | Region of wine origin                          |
16 - region_2           | sub-region of wine origin                      |
17 - variety            | Type of wine                                   |
18 - winery             | Name of the winery                             |



## Load dataset
Read data
```{r}
#Read data
raw <- read.csv('./data/wine-with-type-location.csv', header=TRUE, sep = ";")
```

Add rowid
```{r}
#add rowid
raw <- tibble::rowid_to_column(raw, "ROWID")
head(raw, 5)
# summary(raw)

#numeric data,  'quality' variable is dependent variable 
data <- raw[c(1:12)]
summary(data)
```

## Structure of data
No NAs or Nans or blanks found
```{r}
str(data)
nrow(data)
```

## Visualizations
```{r}
#Melt data
melt_data = melt(data, id.vars=c("ROWID"))

#visualize spread of data
ggplot(melt_data,  mapping = aes(x = value)) + geom_bar(fill = "#FF6666") + facet_wrap(~variable, scales = 'free_x')
boxplot(data[,-c(1)])
```

Above shows data needs scaling and has outliers

## Checking outlier
```{r}
#3 outlier
rosnerTest(data$volatile.acidity, k = 4, warn = F)

#0 outliers
rosnerTest(data$citric.acid, k = 4, warn = F)

#4 outliers
rosnerTest(data$residual.sugar, k = 4, warn = F)

#4 outliers
rosnerTest(data$chlorides, k = 4, warn = F)

#4 outliers
rosnerTest(data$free.sulfur.dioxide, k = 4, warn = F)

#0 outliers
rosnerTest(data$density, k = 4, warn = F)

#2 outliers
rosnerTest(data$pH, k = 4, warn = F)

#4 outliers
rosnerTest(data$sulphates, k = 4, warn = F)

#1 outlier
rosnerTest(data$alcohol, k = 4, warn = F)

```

## Correlations
```{r}
#corr of data
corrmatrix <- cor(data)
corrplot(corrmatrix, method = 'pie', type="upper")
```
Moderate corr of 0.6ish for following
  1) fixed acidity - citric acid
  2) fixed acidity - density
  3) free.sulfur.dioxide - total.sulfur.dioxide

## Removing fixed acidity and total.sulfur
```{r}
#we will remove fixed acidity and total.sulfur
data <- data[-c(1,7)]

#above are visualized and validated via plot
plot(data)
```


# Data preparation
We replaced outliers with 5th and 95th percentile values.

```{r}
#replace outliers with 5th and 95th percentile values
#remember An outlier is not any point over the 95th percentile 
#or below the 5th percentile. Instead, an outlier is considered so 
#if it is below the first quartile – 1.5·IQR or above third quartile + 1.5·IQR.

capOutlier <- function(x){
   qnt <- quantile(x, probs=c(.25, .75), na.rm = T)
   caps <- quantile(x, probs=c(.05, .95), na.rm = T)
   H <- 1.5 * IQR(x, na.rm = T)
   x[x < (qnt[1] - H)] <- caps[1]
   x[x > (qnt[2] + H)] <- caps[2]
   return(x)
}
data$volatile.acidity=capOutlier(data$volatile.acidity)
data$residual.sugar=capOutlier(data$residual.sugar)
data$chlorides=capOutlier(data$chlorides)
data$total.sulfur.dioxide=capOutlier(data$total.sulfur.dioxide)
data$density=capOutlier(data$density)
data$pH=capOutlier(data$pH)
data$sulphates=capOutlier(data$sulphates)
data$alcohol=capOutlier(data$alcohol)

#scale data
data_scaled <- data.frame(scale(data))

#visualize how well data was scaled
boxplot(data_scaled)
#scaling looks good


```

```{r}
silhouette_score <- function(k){
  km <- kmeans(data, centers = k, nstart=25)
  ss <- silhouette(km$cluster, dist(data_scaled))
  mean(ss[, 3])
}
k <- 2:10
avg_sil <- sapply(k, silhouette_score)
plot(k, type='b', avg_sil, xlab='Number of clusters', ylab='Average Silhouette Scores', frame=FALSE)
```

## Show Optimal number of clusters:
```{r}
fviz_nbclust(data_scaled, kmeans, method='silhouette')
```


# PCA and reduction

```{r}

# ##### Build pca using princomp
data_pca1 <- princomp(data_scaled)

#examine the importance of PCs
summary(data_pca1)

# inspect principal components
# loadings shows variance and and how much each variable contributes to each components
loadings(data_pca1)
fviz_pca_var(data_pca1,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```

Above shows that first 8 give us 89% of the variance
We don't see one variable being overbearing

```{r}
#plot
plot(data_pca1)
# scree plot
plot(data_pca1, type = "lines")
#biplot
biplot(data_pca1, col = c("gray", "black"))
```

```{r}
# ###### using princomp first 10 give us 90% of the variance

# ##### build pca using prcomp
data_pca2 <- prcomp(data_scaled)
summary(data_pca2)
#above shows first 6 account for 88% variance

#plot
plot(data_pca2)
# scree plot
plot(data_pca2, type = "lines")
#biplot
biplot(data_pca2, col = c("gray", "black"))

fviz_pca_var(data_pca2,
             col.var = "contrib", # Color by contributions
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```
Above shows first 7 account for 90% variance

Please note: PCA didnt really didnt help at all because of the number of components. We were hoping for a reduction down to 3 - 5 components. stick with scaled data


# Clustering using kmeans

## determine number of Clusters

```{r}
datatocluster <- data_scaled
```

Silhouette method recommends 2
```{r}
fviz_nbclust(datatocluster, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
```

Elbow method seems to show elbow at 4 or 5 
```{r}
fviz_nbclust(datatocluster, kmeans, method = "wss")+
  labs(subtitle = "Elbow method")
```

Gap statistic recommends 1
```{r}
#gap stat takes some time, uncomment to run it

set.seed(123)
#fviz_nbclust(datatocluster, kmeans, nstart = 25,  method = "gap_stat", nboot = 50) + labs(subtitle = "Gap statistic method")

#euclidean takes some time, uncomment to run it
#lets also try nbclust using hierarchical clustering
#recommends 4
#NbClust(data = datatocluster, diss = NULL, distance = "euclidean",min.nc = 2, max.nc = 15, method = "ward.D")
```

Optimal clusters is 4
```{r}
k=4;
```

## kmeans
```{r}
fit <- kmeans(datatocluster, k, nstart=25, iter.max=200)
fit
# Cluster sizes
sort(table(fit$cluster))
```

## visualize and analyze clusters generated by kmeans
```{r}
datatovisualize1 <- data_scaled

#clusplot below, but is not useful as we performed analysis in 9 dimensions
#clusplot below uses first 2 dimensions which covers 47% variability
#too much overlap in 2d
#should really plot pairs or each variable against cluster
clusplot(datatovisualize1, fit$cluster,cex=1,xlab=colnames(datatovisualize1)[1],ylab=colnames(datatovisualize1)[2],col.p=fit$cluster,lines=0,labels=1)

#pairwise plot
#pairwise keeps running and crashes
#uncomment if need to run
#pairs(datatovisualize1, col=c("red","blue","green","yellow")[fit$cluster])

#put quality back in data
datatovisualize1$quality <- raw$quality

old.par <- par(mar = c(2, 2, 2, 2))
par(mfrow=c(2,2))


for(i in 1:10){
  boxplot(datatovisualize1[,i] ~ fit$cluster,
        xlab='Cluster Number', ylab=colnames(datatovisualize1)[i],
        main=paste('Clusters of ', as.character(colnames(datatovisualize1)[i])))
}
#reset graphics
par(old.par)

#cluster 3 is higher quality than others
#why is that?
# cluster 3: alcohol, pH are higher
# cluster 3: sugar, chlorides, density are lower

#put cluster back into raw data so we can save raw data and show in shiny app
raw$cluster_kmeans=fit$cluster

```


# Clustering using Hierarchical Clustering
We implement a Ward’s hierarchical clustering procedure:
```{r}
#distance matrix
d <- dist(data_scaled, method = "euclidean") 
#clustering
h_clust <- hclust(d, method = "ward.D2") 
```

Display dendrogram with cut rects
```{r}
plot(h_clust)
rect.hclust(h_clust , k = 5, border = 2:8)
abline(h = 5, col = 'red')
```

Display clusters in different colors
```{r}
avg_dend_obj <- as.dendrogram(h_clust)
avg_col_dend <- color_branches(avg_dend_obj, h =  40)
plot(avg_col_dend)
```

Extract clusters
```{r}
groups <- cutree(h_clust,k=5)
```

Count number of instances in each group
```{r}
data_df_cl <- mutate(data, cluster = groups)
count(data_df_cl, cluster)
```

We can analysis the quality and
```{r}
summary(data_df_cl)

ggplot(data_df_cl, aes(x=density, y = residual.sugar , color = factor(cluster))) + geom_point()
clusplot(data_df_cl, fit$cluster,cex=1,xlab=colnames(data_df_cl)[1],ylab=colnames(data_df_cl)[2],col.p=groups,lines=0,labels=1)

```


## visualize and analyze clusters generated by hierarchical clustering
```{r}
datatovisualize2 <- data_scaled
```

2D representation of the Segmentation:
```{r}
clusplot(datatovisualize2, groups, main='Clusters')
```

```{r}
# pairwise plot
# pairs(datatovisualize2, col=c("red","blue","green","yellow")[groups])
```

```{r}
#put quality back in data
datatovisualize2$quality <- raw$quality
par(old.par)
par(mfrow=c(2,2))

for(i in 1:10){
  boxplot(datatovisualize2[,i] ~ groups,
        xlab='Cluster Number', ylab=colnames(datatovisualize2)[i],
        main=paste('Clusters of ', as.character(colnames(datatovisualize2)[i])))
}
#reset graphics
par(old.par)

#cluster 4 is higher quality than others
#why is that?
# cluster 4: alcohol is higher
# cluster 4: chlorides, density are lower

#put cluster back into raw data so we can save raw data and show in shiny app
raw$cluster_hclust=groups
#TODO save data to file to load in shiny


save(raw, file = "./WineAnalysis/shiny.RData")

```


# Shiny App
Link: [https://yumingcui.shinyapps.io/WineAnalysis/]
